prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments

### 2022-03-18 inst/MyArticle/paper.Rmd
Cosmetic text edits

Label-Free Quantification (LFQ) Proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment vs. control are used [@de2022apoe2,@laubscher2021baf]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic,@meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested.  Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [TODO ...] or validation experiments.

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in _prolfqua_ if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in _MSStats_ format, only a few lines of code to update the _prolfqua_ configuration are needed. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable (e.g., intensity) in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enabled us to easily interface with many data manipulation, visualization, and modeling methods implemented in base __R__[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported (e.g. auto-completion) and easy to use (OOP).

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with _prolfqua_ in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `prolfqua` (see Figure \@ref(fig:ContrastUML)).

@@

The peptide intensities from the _peptide.txt_ file were $log_2$ transformed, and subsequent robust z-score transformation, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e. proteins with a single peptide assignment. No further data filtering was used. We also did not use imputation. The protein level linear model was fitted to the protein intensities inferred from the peptide intensities using the Tukey's median polish. The mixed effect linear model, and the `r BiocStyle::Biocpkg("PECA")` model were fitted to the peptide intensities. The details of the _prolfqua_ benchmark implementation are available online  [BenchmarkingIonstarData](https://wolski.github.io/prolfqua/articles/BenchmarkingIonstarData.html).

@@

The Benchmark functionality of _prolfqua_ includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g. the false discovery proportion FDP. We use it to study how well linear, mixed effect models or $p$-value moderation model quantitative mass spectrometric high throughput experiments and compare it with results produced by MSstats [@MSstats2014] and by proDA [@bioxrvproDA2020]. Figure \@ref(fig:FDRfdp)

The IonStar  [@shen2018ionstar] benchmark dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentrations. We know that for _H. sapiens_ proteins the difference $\beta$ between two dilutions should be $\beta = 0$ while for _E. coli_ proteins, we know that the difference between dilutions should be $\beta \ne 0$. To compare the performance of the various methods implemented in _prolfqua_ we use only the contrasts resulting in small fold-changes $\beta = 1.2,1.25,1.3(3),1.5$.

@@

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:FDRfdp) right panel draws on the horizontal axis the FDR determined from the model, and on the vertical axis, the FDP obtained from the confusion matrix.  We see that most of the lines are below the diagonal, indicating that the FDR estimates obtained with _prolfqua_ are modestly conservative for this benchmark dataset provide. In the case of _MSstats_, we observe a high proportion of false discoveries for small $FDR$ values. In the case of `r BiocStyle::Biocpkg("PECA")`, the FDR estimates, obtained when Benjamini-Hochberg correcting the regulation probabilities, strongly overestimate the $FDP$.

@@

The probabilistic dropout analysis implemented in the _proDA_ produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled $p$-values and $t$-statics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the _proDA_ package as an additional modeling option into _prolfqua_ (see Figure \@ref(fig:ContrastUML)).

@@

_prolfqua_ allows for considerable flexibility to model your experiments. Various types of models are available, and the contrast specification is consistent and intuitive for all models. The benchmarking results are comparable with other methods. The modular design of _prolfqua_, allows for customizing the analysis or adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), quickly.

Using the tidy table to model the data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., _protti_. Furthermore, the use of `R6` classes, which encapsulate the configuration and the data, allows writing very concise code (no function arguments needed), and auto-completion support in the Rstudio editor makes it easy for novices to explore _prolfquas_ functionality.

In summary _prolfqua_ is an easy-to-use and feature rich R package to analyse quantitative mass spectrometric data, report results, and benchmark MS software and statistical methods. We provide more details at the website https://wolski.github.io/prolfqua/. To install:

@@

|       |                                                   |
|:------|:--------------------------------------------------|
|AUC    |Area Under the Curve                               |
|ESI-MS |Electro-Spray-Ionization Mass Spectrometry         |
|FDR    |False Discovery Rate                               |
|HPLC   |High-Pressure Liquid Chromatography                |
|LC-MS  |Liquid Chromatography followed by Mass Spectrometry|
|LFQ    |Label-Free Quantification                          |
|LOD    |Limit Of Detection                                 |
|MS     |mass spectrometry                                  |
|OO     |Object-Oriented                                    |

>>>>

Label-Free Quantification (LFQ) Proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment vs. control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested.  Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [TODO ...] or validation experiments.

@@

We store all the data needed for analysis in a single data frame in a tidy table, i.e., every column is a variable, every row is an observation, every cell is a single value[@tidydata2014]. Using an __R6__[@R6cite] configuration object, we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy _Spectronaut_, or _Skyline_ outputs, or data in _MSStats_ format, only a few lines of code to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` configuration are needed. For popular software like _MaxQuant_ or _MSFragger_, which stores the same variable (e.g., intensity) in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enabled us to easily interface with many data manipulation, visualization, and modeling methods implemented in base __R__[@RCoreTeam2021] and the tidyverse[@tidyverse2019]. We use __R6__ classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). __R6__ classes are well supported (e.g. auto-completion) and easy to use (OOP).

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `prolfqua` (see Figure \@ref(fig:ContrastUML)).

@@

The peptide intensities from the _peptide.txt_ file were $log_2$ transformed, and subsequent robust z-score transformation, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e. proteins with a single peptide assignment. No further data filtering was used. We also did not use imputation. The protein level linear model was fitted to the protein intensities inferred from the peptide intensities using the Tukey's median polish. The mixed effect linear model, and the `r BiocStyle::Biocpkg("PECA")` model were fitted to the peptide intensities. The details of the `r BiocStyle::Githubpkg("fgcz/prolfqua")` benchmark implementation are available online  [BenchmarkingIonstarData](https://wolski.github.io/prolfqua/articles/BenchmarkingIonstarData.html).

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g. the false discovery proportion FDP. We use it to study how well linear, mixed effect models or $p$-value moderation model quantitative mass spectrometric high throughput experiments and compare it with results produced by MSstats [@MSstats2014] and by proDA [@bioxrvproDA2020]. Figure \@ref(fig:FDRfdp)

The IonStar  [@shen2018ionstar] benchmark dataset contains _H. sapiens_ proteins with constant concentrations and _E. coli_ proteins with varying concentrations. We know that for _H. sapiens_ proteins the difference $\beta$ between two dilutions should be $\beta = 0$ while for _E. coli_ proteins, we know that the difference between dilutions should be $\beta \ne 0$. To compare the performance of the various methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` we use only the contrasts resulting in small fold-changes $\beta = 1.2,1.25,1.3(3),1.5$.

@@

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:FDRfdp) right panel draws on the horizontal axis the FDR determined from the model, and on the vertical axis, the FDP obtained from the confusion matrix.  We see that most of the lines are below the diagonal, indicating that the FDR estimates obtained with `r BiocStyle::Githubpkg("fgcz/prolfqua")` are modestly conservative for this benchmark dataset provide. In the case of _MSstats_, we observe a high proportion of false discoveries for small $FDR$ values. In the case of `r BiocStyle::Biocpkg("PECA")`, the FDR estimates, obtained when Benjamini-Hochberg correcting the regulation probabilities, strongly overestimate the $FDP$.

@@

The probabilistic dropout analysis implemented in the _proDA_ produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled $p$-values and $t$-statics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the _proDA_ package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model your experiments. Various types of models are available, and the contrast specification is consistent and intuitive for all models. The benchmarking results are comparable with other methods. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for customizing the analysis or adding new features, e.g., generalized linear models (_glm_'s) for missing data modeling or robust linear models (_rlm_'s), quickly.

Using the tidy table to model the data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")`. Furthermore, the use of `R6` classes, which encapsulate the configuration and the data, allows writing very concise code (no function arguments needed), and auto-completion support in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")`s functionality.

In summary `r BiocStyle::Githubpkg("fgcz/prolfqua")` is an easy-to-use and feature rich R package to analyse quantitative mass spectrometric data, report results, and benchmark MS software and statistical methods. We provide more details at the website https://wolski.github.io/prolfqua/. To install:

@@

|       |                                                   |
|:------|:--------------------------------------------------|
|AUC    |Area Under the Curve                               |
|ESI-MS |Electro-Spray-Ionization Mass Spectrometry         |
|FDR    |False Discovery Rate                               |
|HPLC   |High-Pressure Liquid Chromatography                |
|LC-MS  |Liquid Chromatography followed by Mass Spectrometry|
|LFQ    |Label-Free Quantification                          |
|LOD    |Limit Of Detection                                 |
|MS     |mass spectrometry                                  |
|OO     |Object-Oriented                                    |
|UML    |Unified Modeling Language                          |
